---
title: "Survival Analysis of Breast Cancer Cases: Clustering, Choosing the Value of k"
author: "Dr. Nishat Mohammad"
date: "2024-07-25"
output:
  word_document:
    toc: yes
  pdf_document:
    latex_engine: xelatex
  html_document:
    toc: yes
    toc_float: yes
    code_folding: show
---


## 7. K-Means Clustering. 

```{r kmeans_clustering}
# Scale expression data using z-score method
data_expression_scaled <- scale(data_expression2)

# Add row names
rownames(data_expression_scaled) <- data_merged$PATIENT_ID

# Get the top 100 gene expressions
data_expression_top100 <- data_expression_scaled[, top_genes$Gene]
data_expression_top100[1:10,1:10]

```
Here the expression data for the top 100 genes was extracted.  
 
 
### 7.1. Number of Clusters.  

```{r cluster_number}
# Range for clusters
k_range <- 2:10

# Within-cluster sum of squares (WCSS) for each k value
wcss <- vector("numeric", length(k_range))

# k-means for each value of k and get WCSS
for (i in seq_along(k_range)) {
  k <- k_range[i]
  kmeans_res <- kmeans(data_expression_top100, centers = k, nstart = 25)
  wcss[i] <- kmeans_res$tot.withinss
}

# Elbow plot
fviz_nbclust(data_expression_top100, kmeans, method = "wss", k.max = 10, nstart = 25)

# silhouette method
fviz_nbclust(data_expression_top100, kmeans, method = "silhouette")+
  labs(subtitle = "Silhouette method")

# Gap_Stat is very slow
#set.seed(786)
#fviz_nbclust(data_expression_top100, kmeans, nstart = 25,  method = "gap_stat", nboot = 500, verbose=FALSE)+
#  labs(subtitle = "Gap statistic method")

# Using NbClust()
set.seed(786)  # For reproducibility
nbclust_res <- NbClust(data = data_expression_top100, 
                           distance = "euclidean", 
                           min.nc = 2, 
                           max.nc = 10, 
                           method = "kmeans")


# Correlation matrix for the distance
dist.cor <- get_dist(data_expression_top100, method = "spearman")
fviz_dist(dist.cor)
nbclust_spearman <- NbClust(data = data_expression_top100,
        diss = dist.cor,
        distance = NULL,
        min.nc = 2,
        max.nc = 10,
        method = "kmeans")


```
The elbow and Silhouette Plots were not conclusive. Using `NbClust()`, the Spearman correlation was applied to the `diss` option with NULL added to the distance option. Finally, 2 as the value of k number of clusters based on the Hubert Statistics Plots.  


### 7.2 Clustering with k=2.    

```{r Kmeans_k2}
set.seed(786)
kmeans_res <- kmeans(data_expression_top100, centers = 2, nstart = 25)
str(kmeans_res)

# Visualize clusters
fviz_cluster(kmeans_res, data = data_expression_top100)

```

Note the two clusters are overlapping, higher values of K will be tried out.

### 7.3. K-means clustering with higher values of k.  

```{r Increase_k}
k3 <- kmeans(data_expression_top100, centers = 3, nstart = 25)
k4 <- kmeans(data_expression_top100, centers = 4, nstart = 25)
k5 <- kmeans(data_expression_top100, centers = 5, nstart = 25)

# plots to compare
p1 <- fviz_cluster(kmeans_res, geom = "point", data = data_expression_top100) + ggtitle("k = 2")
p2 <- fviz_cluster(k3, geom = "point",  data = data_expression_top100) + ggtitle("k = 3")
p3 <- fviz_cluster(k4, geom = "point",  data = data_expression_top100) + ggtitle("k = 4")
p4 <- fviz_cluster(k5, geom = "point",  data = data_expression_top100) + ggtitle("k = 5")

grid.arrange(p1, p2, p3, p4, nrow = 2)

# Final Kmeans with k = 2
final_kmeans <- kmeans_res
fviz_cluster(final_kmeans, geom = "point",  data = data_expression_top100) + ggtitle("k = 2")

# Perform t-SNE
tsne_res <- Rtsne(as.matrix(data_expression_top100), dims = 2, perplexity = 30)
# t-SNE result data frame
tsne_df <- as.data.frame(tsne_res$Y)
tsne_df$Cluster <- as.factor(final_kmeans$cluster)
# Visualize t-SNE result
ggplot(tsne_df, aes(x = V1, y = V2, color = Cluster)) +
  geom_point() +
  ggtitle("t-SNE Plot of Clusters") +
  theme_minimal()


# Perform UMAP
umap_res <- umap(data_expression_top100)

# UMAP result in a data frame
umap_df <- as.data.frame(umap_res$layout)
umap_df$Cluster <- as.factor(final_kmeans$cluster)

# Visualize UMAP result
ggplot(umap_df, aes(x = V1, y = V2, color = Cluster)) +
  geom_point() +
  ggtitle("UMAP Plot of Clusters") +
  theme_minimal()

```

The directive from business specifies the use of kmeans for clustering and thus has to be followed through but other methods of clustering may be used to look for better outlined clusters. For now, we move forward with kmeans using 2 clusters. The clusters can be seen using tSNE and UMAP.  

### 7.4. Assign clusters to data.  

```{r assign_clusters}
data_expression_clustered <- data.frame(data_expression_top100,final_kmeans$cluster)
# Samples in each cluster
print(table(data_expression_clustered$final_kmeans.cluster))

```
Above are the number of patients in cluster 1 and 2 being 346 and 581 respectively.  
